Properties of winning Iterated Prisoner's Dilemma strategies.

### 2020-03-27 paper/main.tex
finalise introduction

%%%%Packages%%%%
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{graphics}
\usepackage{multicol}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{hyperref}
\usepackage{color,colortbl}
\usepackage{array}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{wrapfig, blindtext}
\usepackage[table]{xcolor}
%%%%%%%%%%%%%%%%%

@@

\begin{abstract}
From the celebrated performance of Tit for Tat, to the
introduction of the zero-determinant strategies, to the use of sophisticated
structures such as neural networks, the literature has been exploring the
performance of Iterated Prisoner's Dilemma strategies for decades. The results of the literature,
however, have been relying on a small number of somewhat arbitrarily selected
strategies in a very small number of tournaments,
casting doubt on the generalizability of results from statistical perspective.
This manuscript evaluates \numberofstrategies strategies in 45686
tournaments, presents the top performing strategies, and analyzes their
salient features. The results of the tournaments imply that there is not a single
type of strategy that performs well in any Iterated Prisoner's Dilemma interaction.
There are several properties, however, that heavily influence the best performing
strategies: be nice, be provocable, be a little envious and be clever.
\end{abstract}

@@

In the 80's, Axelrod performed two computer tournaments~\cite{Axelrod1980a,
Axelrod1980b}. The
winner of both tournaments was the simple strategy Tit For Tat which cooperated
on the first turn and then simply copied the previous action of its opponent.
Axelrod concluded that the strategy's robustness was due to four different
properties:

\begin{itemize}
    \item Do not be envious
    \item Be "nice"; Do not be the first to defect
    \item Reciprocate both cooperation and defection
    \item Do not be too clever
\end{itemize}

@@

Further research proved that the strategy had weakness, and more
specifically, it was shown that the strategy suffered in environments with
noise~\cite{Bendor1991, Donninger1986, Molander1985, Hammerstein1984}. This was
mainly due to the strategy's lack of generosity and contrition. The strategy was
quick to punish a defection, and in a noisy environment it could lead to a
repeated cycle of defections and cooperations. Some new strategies, more
robust in tournaments with noise, were soon introduced and became the new
protagonists of the game. These include Nice and Forgiving~\cite{Bendor1991},
Pavlov~\cite{Nowak1993} and Generous Tit For Tat~\cite{Nowak1992}.
Several works extended the reciprocity based approach which has led to new
strategies. For example Gradual~\cite{Beaufils1997} which was constructed to
have the same qualities as those of Tit for Tat except one,
\textit{Gradual} had a memory of the game since the beginning of it. Gradual
recorded the number of defections by the opponent and punished them with a
growing number of defections. It would then enter a calming state in which it
would cooperates for two rounds. In a tournament of 12 strategies, including
both Tit for Tat and Pavlov, Gradual managed to outperformed them all. A
strategy with the same intuition as Gradual is \textit{Adaptive Tit for
Tat}~\cite{tzafestas-2000a}. Adaptive Tit for Tat does not keep a permanent
count of past defections, it maintains a continually updated estimate of the
opponent’s behaviour, and uses this estimate to condition its future actions. In
the exact same tournament as in~\cite{Beaufils1997} with now 13 strategies Adaptive
Tit for Tat ranked first.
A set of IPD strategies that have received a lot of attention
are the zero-determinant strategies (ZDs)~\cite{Press2012}. By
forcing a linear relationship between the payoffs ZDs can ensure that they will
never receive less than their opponents. The
American Mathematical Society's news section stated that ``the world of game
theory is currently on fire''.
ZDs are indeed a set of mathematically unique strategies
and robust in pairwise interactions, however, their superiority in tournament settings
have been tested. In~\cite{Harper2017} a tournament containing over
200 strategies, including ZDs, was ran and none of them
ranked in top spots. Instead, the top ranked strategies were a set of
trained strategies based on lookup tables~\cite{Axelrod1987}, hidden markov
models~\cite{Harper2017} and finite state automata~\cite{Miller1996}.

Though only select pieces of work have been discussed, the IPD literature is
rich, and new strategies and competitions are being published every year~\cite{Glynatsi2019}. The
question, however, still remains the same: what is the best way to play the
game? Compared to other works, whereas a few selected strategies are evaluated
on a small number of tournaments, this manuscript evaluates the performance of \numberofstrategies
strategies in \numberofalltournaments tournaments. These tournaments come in a
number of variations including tournaments with noise, probabilistic match
length and both noise and probabilistic match length.
The later part of the paper, evaluates
the impact of features on the performance of the strategies using modern
machine learning techniques. These features include measures regarding a
strategy's behaviour and measures regarding the tournaments. The outcomes of our
work reinforce the discussion started by Axelrod:

\begin{itemize}
    \item Do not be envious
    \item Be "nice"; Do not be the first to defect
    \item Reciprocate both cooperation and defection
    \item Do not be too clever
\end{itemize}

Furthermore, the data set used
in this work has been made publicly available~\cite{data} and can be used
for further analysis and insights.

The different tournament types as well as the data collection, which is made
possible due to an open source package called Axelrod-Python,
are covered in Section~\ref{section:data_collection}.
Section~\ref{section:top_performances}, focuses on the best performing
strategies for each type of tournament and overall.
Section~\ref{section:evaluation_of_performance}, explores the traits which
contribute to good performance, and finally the results are summarised in
Section~\ref{section:conclusion}. This manuscripts uses several parameters.
These are introduced in the following sections, however, the full set of
parameters and their definitions are given in Appendix~\ref{app:parameters}.

>>>>

%%%%Packages%%%%
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{graphics}
\usepackage{multicol}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{hyperref}
\usepackage{color,colortbl}
\usepackage{array}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{wrapfig, blindtext}
\usepackage{soul}
\usepackage[table]{xcolor}
%%%%%%%%%%%%%%%%%

@@

\begin{abstract}
From the celebrated performance of Tit for Tat, to the
introduction of the zero-determinant strategies, to the use of sophisticated
structures such as neural networks, the literature has been exploring the
performance of Iterated Prisoner's Dilemma strategies for decades. The results of the literature,
however, have been relying on a small number of somewhat arbitrarily selected
strategies in a very small number of tournaments,
casting doubt on the generalizability of results from statistical perspective.
This manuscript evaluates \numberofstrategies strategies in 45686
tournaments, presents the top performing strategies, and analyzes their
salient features. The results imply that there is not a single
strategy that performs well in any Iterated Prisoner's Dilemma interaction.
There are several properties, however, that heavily influence the best performing
strategies, these are: be nice, be provocable, be a little envious, be clever
and adapt to the environment.
\end{abstract}

@@

Axelrod performed two computer tournaments~\cite{Axelrod1980a,
Axelrod1980b} in the 80's. The
winner of both tournaments was the simple strategy Tit For Tat which cooperated
on the first turn and then simply copied the previous action of its opponent.
Axelrod concluded that the strategy's robustness was due to four properties, which
he adapted in four suggestions on doing well in an IPD:

\begin{itemize}
    \item Do not be envious
    \item Be "nice"; Do not be the first to defect
    \item Reciprocate both cooperation and defection; Be provocable
    \item Do not be too clever
\end{itemize}

@@

There are strategies which have built upon Tit For Tat, and the reciprocity based approach.
In~\cite{Beaufils1997} Gradual was introduced which was constructed to have the
same qualities as those of Tit for Tat except one, Gradual had a memory
of the game since the beginning of it. Gradual recorded the number of defections
by the opponent and punished them with a growing number of defections. It would
then enter a calming state in which it would cooperates for two rounds. A
strategy with the same intuition as Gradual is Adaptive Tit for
Tat~\cite{tzafestas-2000a}. Adaptive Tit for Tat does not keep a permanent
count of past defections, it maintains a continually updated estimate of the
opponent's behaviour, and uses this estimate to condition its future actions.

Other works have built upon the limitations of the strategy, and others have shown
that suggestions made by Axelrod were incomplete.
In~\cite{Bendor1991, Donninger1986, Molander1985, Hammerstein1984} it was shown
that Tit For Tat suffered in environments with noise. This was
mainly due to the strategy's lack of generosity and contrition. The strategy was
quick to punish a defection, and in a noisy environment it could lead to a
repeated cycle of defections and cooperations. Some new strategies, more
robust in tournaments with noise, were soon introduced and became the new
protagonists of the game. These include Nice and Forgiving~\cite{Bendor1991},
Pavlov~\cite{Nowak1993} and Generous Tit For Tat~\cite{Nowak1992}.
In~\cite{Press2012} a set of envious IPD strategies that have received a lot of
attention were introduced called the zero-determinant strategies (ZDs). By
forcing a linear relationship between the payoffs ZDs can ensure that they will
never receive less than their opponents. ZDs are indeed a set of mathematically
unique strategies and robust in pairwise interactions, however, their
superiority in tournament settings have been tested. In~\cite{Harper2017} a
series of clever strategies were introduced, and a tournament containing over
200 strategies was ran and none of the aforementioned strategies ranked in top
spots. Instead, the top ranked strategies were the set of clever trained
strategies based on lookup tables~\cite{Axelrod1987}, hidden markov
models~\cite{Harper2017} and finite state automata~\cite{Miller1996}.

Though only select pieces of work have been discussed, there is a board number
of strategies in the literature, and new strategies and competitions are being
published every year~\cite{Glynatsi2019}. The question, however, still remains
the same: what is the best way to play the game?

Compared to other works, whereas a few selected strategies are evaluated
on a small number of tournaments, this manuscript evaluates the performance of \numberofstrategies
strategies in \numberofalltournaments tournaments. These tournaments come in a
number of variations including tournaments with noise, probabilistic match
length and both noise and probabilistic match length.
The later part of the paper, evaluates
the impact of features on the performance of the strategies using modern
machine learning techniques. These features include measures regarding a
strategy's behaviour and measures regarding the tournaments. The outcomes of our
work reinforce the discussion started by Axelrod, and it concludes that the properties
of a successful strategy in the IPD are:

\begin{itemize}
    \item \st{Do not be envious}
    \item Be "nice"; Do not be the first to defect
    \item Reciprocate both cooperation and defection; Be provocable
    \item \st{Do not be too clever}
    \item Adapt to the environment; Adjust to the mean cooperator
\end{itemize}

The different tournament types as well as the data collection, which is made
possible due to an open source package called Axelrod-Python,
are covered in Section~\ref{section:data_collection}. The data set generated
for this work has been made publicly available~\cite{data} and can be used
for further analysis and insights.
Section~\ref{section:top_performances}, focuses on the best performing
strategies for each type of tournament and overall.
Section~\ref{section:evaluation_of_performance}, explores the traits which
contribute to good performance, and finally the results are summarised in
Section~\ref{section:conclusion}. This manuscripts uses several parameters.
These are introduced in the following sections, however, the full set of
parameters and their definitions are given in Appendix~\ref{app:parameters}.
